---
title: "Longitudinal Binary Model"
output: html_document
---

```{r setup, include=FALSE}
library(dplyr)
```
## Description of Data and Research Question

The data comes from user fitbit data and includes variables such as step count, time spent in different activity levels, and sleep. The user is identified by a unique ID and is repeatedly observed across time (in days). The research question is:

> to determine if the number of steps taken by a user is related to whether or not they slept more than 7 hours that night.

This is a longitudinal binary model because the response variable is binary (slept more than 7 hours or not) and the data is longitudinal (repeated observations across time). The prediction will be a probability of sleeping more than 7 hours.

We will include a random intercept for each individual to account for the individual's "baseline" probability of sleeping more than 7 hours.

### Data Exploration

```{r}
activity_data <- read.csv("data/Fitabase Data 4.12.16-5.12.16/dailyActivity_merged.csv")
sleep_data <- read.csv("data/Fitabase Data 4.12.16-5.12.16/sleepDay_merged.csv")

activity_data$ActivityDate <- as.Date(activity_data$ActivityDate, format = "%m/%d/%Y")

# calculate the sleep binary for each day
sleep_data$sleep_binary <- ifelse(sleep_data$TotalMinutesAsleep > 7*60, 1, 0)
# cast the SleepDay to date from datetime and name ActivityDate
# first from char to datetime
sleep_data$SleepDay <- as.POSIXct(sleep_data$SleepDay, format = "%m/%d/%Y %I:%M:%S %p")
# then from datetime to date
sleep_data$ActivityDate <- as.Date(sleep_data$SleepDay)

print(str(activity_data))
print(str(sleep_data))
```

We join the data. 

```{r}
# join the data on ActivityDate and Id
data <- merge(
    activity_data,
    sleep_data,
    by = c("Id", "ActivityDate")
)

print(str(data))
```

We display the count of observations for each idividual.

```{r}
data %>%
    group_by(Id) %>%
    summarise(n = n()) %>%
    arrange(desc(n))
```

We see that we do not have a uniform numer of observations for each individual. 

## Model Setup

### Random Intercept Model

Random intercept model. For subject i’s jth observation, the random intercept model is defined as:

$$y_{ij} = \beta_0 + \beta_1x_{ij} + u_i + \epsilon_{ij}$$

where:

* $y_ij$ is the probability of sleeping more than 7 hours for subject i on day j
* $x_{ij}$ is the number of steps taken by subject i on day j
* $u_i$ is the random intercept for subject i
* $\epsilon_{ij} \sim N(0,\sigma_e^2)$ is the error term for subject i on day j

### Porbit Model

In our case, we follow Albert and Chib (1993, Note set II page 160) to represent bernoulli data as a latent variable model (equivalent to a probit model in the GLM frameworks).

We let $z_i$ denote subjet i's latent variable:

$$y_i = \begin{cases} 1 & \text{if } z_i > 0 \\ 0 & \text{if } z_i \leq 0 \end{cases}$$

where $z_i \sim N(x_i^T\beta, 1)$ which is never observed. 

The likelihood for a single subject $i$ is then:

$$\mathbb{L}(y_i|z_i,x_i,\beta)\propto 
\{1(z_i>0)1(y_i=1) + 1(z_i \leq 0)1(y_i=0)\} 
\exp{\left(
-\frac{1}{2}(z_i - x_i^T\beta)^2
\right)}$$

In our case, unlike in Albert and Chib, we do not only have the $\beta$ parameter. We also have the random intercept parameter $u_i$ which we typically assume to have a prior $u_i \sim N(0, \sigma_u^2)$. Our choice of prior for $\beta$ determines the posterior distribution of $\beta$. We recall the bayesian relationship between the posterior and the likelihood:

$$p(\beta|y) \propto \mathbb{L}(y|\beta)\pi(\beta)\pi(u_i)$$.

### Priors

If we choose a non-informative prior $\pi(\beta)$ then we get the following posterior:

$$p(\beta|y) \propto \pi(\beta)\prod_{i=1}^n\mathbb{L}(y_i|z_i,x_i,\beta)$$

We can also choose the prior for $\beta$ to conduct a ridge or lasso regression by choosing $\beta \sim N(0, \gamma)$ or $\beta \sim \text{Laplace}(0, \gamma)$ respectively.

# Logistic Regression (Just for now)

```{r}
# logistic regression using base R
logit <- glm(sleep_binary ~ TotalSteps, data = data, family = binomial(link = "logit"))
summary(logit)
```